---
title: 'R Notebook: #66daysof(health)data - Day Three'
subtitle: Measures of Central Tendency, Power, Confidence Intervals & doing Hypothesis Testing using the Cleveland Heart Attack Data
output:
  html_document:
    df_print: paged
---

```{r}
library(tidyverse)

# Read in Heart Attacks data set
heart <- read_csv(
  "Data/heart.csv", 
  col_types = list(
    age = col_integer(),
    sex = col_number(),
    cp = col_number(),
    trtbps = col_integer(),
    chol = col_integer(),
    fbs = col_logical(),
    restecg = col_number(),
    thalachh = col_integer(),
    exng = col_logical(),
    oldpeak = col_double(),
    slp = col_number(),
    caa = col_integer(),
    thall = col_number(),
    output = col_logical()
  )
)

spec(heart)
```
```{r}
head(heart, 10)
```

```{r}
# Recode Factors 

heart$sex <- as.factor(heart$sex)
levels(heart$sex) <- c('FEMALE', 'MALE')

heart$cp <- factor(heart$cp, levels=c('3', '2', '1', '0'))
levels(heart$cp) <- c('ASYMPT', 'NON-ANG PAIN', 'ATYP ANG', 'TYP ANG')

heart$restecg <- as.factor(heart$restecg)
levels(heart$restecg) <- c('NORM', 'ABN ST-T', 'LVH')

heart$slp <- factor(heart$slp, levels=c('1', '2', '0'))
levels(heart$slp) <- c('FLAT', 'DOWNSLP', 'UPSLP')

# Can't translate this one because I can't find a data dictionary that shows what these codes (0-3) mean
heart$thall <- as.factor(heart$thall)
head(heart, 10)

```

```{r}
# Measures of Central Tendency

summary(heart)
```
```{r}

nrow(heart)
```

Yesterday, we calculated power and did hypothesis testing using Pearson's R correlation and two-sample t-testing.  

Today, let's shift gears a bit, and move in another direction.  Let's do some categorical analysis with Chi-Sq and CMH testing!

First, let's try a Chi-Sq test to see if there may be a difference in relationship between people with exercise-induced angina (exng) and coronary artery disease (output) vs. those without.  

```{r}
# A preliminary plot, to get a look at the data

ggplot(heart) +
  aes(x = exng, fill = output) +
  geom_bar(position = "fill")


```


It looks like the relationship between exercise-induced angina and vessel narrowing isn't what I expected!

```{r}
# Create a table with the needed variables.
heart.data = table(heart$exng, heart$output) 
print(heart.data)

# Perform the Chi-Square test.
print(chisq.test(heart.data))

# Calculate the critical value.
writeLines("The critical value is:")
qchisq(0.05, 1)

```

On the left side of the table, we have Exercise-Induced Angina T/F, and on the top of the table, we have Coronary Artery Disease, T/F.  

Of those **without** exercise-induced angina, 70% had coronary artery disease!  Of those with exercise-induced angina, only 23% had coronary artery disease!  So, while there appears to be a significant effect size (55.95, which is above the critical value of 0.004) and p-value <.0001, the direction of the relationship isn't what I would have expected.

Now, let's calculate the power, with 1 df:


```{r}
library(pwr)

pwr.chisq.test(w=0.2, N=303, df=1, sig.level=0.05)
```

Having 303 rows, we can detect a small (0.2) effect size with 93.6% power at alpha=0.05.

This means that we only have a 6.4% chance of failing to reject the NULL hypothesis if the NULL hypothesis is FALSE (making a Type II error).

And, since we set alpha = 0.05, we have a 5% chance of rejecting the NULL hypothesis when it is actually TRUE (making a Type I error).

```{r}
library(pwr)

pwr.chisq.test(w=0.2, df=1, power=0.80, sig.level=0.05)
```
To achieve 80% power and detect a small effect (0.2), we need at least 197 samples.

Now, let's apply The Cochran-Mantel-Haenszel test by performing the same chi-sq test again, but breaking down the results by sex.

H0: There is NO significant three way interaction between exercise-induced angina, coronary artery disease and sex. There is no significant confounding influence of sex on the relationship between exercise-induced angina and CAD / vessel narrowing.

HA: There IS a significant three way interaction between exercise-induced angina, coronary artery disease and sex. Exercise-induced angina and coronary artery disease are not independent within each level of sex. There IS a significant confounding influence of sex on the relationship between exercise-induced angina and CAD / vessel narrowing.


```{r}

ggplot(heart) +
  aes(x = exng, fill = output) +
  geom_bar(position = "fill") + 
  facet_wrap(~sex)

```

Surprisingly, it looks like females were more likely to have CVD overall (OUTPUT=TRUE) in this data set, but those **without** exercise-induced angina were **much** more likely to have CVD than males without the same.

```{r}
library(vcd)
library(rcompanion)

heart.table <- table(heart$exng, heart$output, heart$sex)
heart.table

heart.cmh <- mantelhaen.test(heart$exng,heart$output,heart$sex, alternative = 'two.sided')
heart.cmh

woolf_test(heart.table)

# post-Hoc
groupwiseCMH(heart.table,
             group   = 3,
             fisher  = TRUE,
             gtest   = FALSE,
             chisq   = FALSE,
             method  = "fdr",
             correct = "none",
             digits  = 3)

```


Proportionally fewer men than women had CAD / vessel narrowing at all in this sample (output=TRUE).

Chi-squared = 44.98, Common odds ratio = 0.141, p-value = <.0001

The effect size (Chi-squared) is smaller here than in the previous chi-sq test, and the common odds ratio is the weighted average of the stratum-specific estimates.  This has the effect of controlling the effect size and OR magnitude of the relationship between exercise-induced angina and CAD for the confounding effect of sex.  

Recall OR = ad/bc where a b c d are the cells in a 2x2 frequency table - 

So, in the original (unadjusted) table, the odds ratio of exercise-induced angina to CAD would have been:

   OR = (62 x 23)/ (142 x 76), or 0.132135
   
Since the odds ratio is < 1, the odds of CAD are 0.132 times LOWER given exercise-induced angina, compared to no exercise-induced angina.

The CMH adjusted OR is 0.141, so once we remove the confounding influence of sex, the odds of CAD are 0.141 times LOWER given exercise-induced angina, compared to no-exercise-induced angina.

The 95% confidence interval for the CMH-adjusted OR indicates that we can be 95% confident that the reduction in CVD given exercise-induced angina is between 0.07945121 and 0.24929225 times.  

The Woolf-test indicates the two data sets are comparable.

The CMH result indicates that there is a significant difference in association between exercise-induced angina and cardiovascular disease (vessel narrowing) when data are stratified by sex.  

The post-Hoc test uses Fisher's Exact Test to demonstrate that p-values are significant for a relationship between given and expected counts of exercise-induced angina x CAD for both females and males.




